Concurrent Programming with Harmony

### 2020-07-20 book/paper.tex
Added some stuff
Improved interrupt chapter

Processes can be \emph{interrupted}.  An interrupt is a notification of some event
such as a keystroke, a timer expiring, the reception of a network packet, the completion
of a disk operation, and so on.  We distinguish \emph{interrupts} and \emph{exceptions}.
An exception is caused by the process executing an invalid machine instruction such as
divide-by-zero.  An interrupt is caused by some peripheral device.
\index{exception}

Harmony allows modeling interrupts using operator \texttt{trap}:
\begin{code}
\textbf{trap} \textit{handler} \textit{argument}
\end{code}
invokes \textit{handler argument} at some later, unspecified time.
Only one of these asynchronous events can be outstanding at a time; a new call to
\textbf{trap} overwrites any outstanding one.
\autoref{fig:trap} gives an example of how \textbf{trap} might be used.
Here, the \textit{main}() process loops until the interrupt has occurred and
the \textit{done} flag has been set.

But now consider \autoref{fig:trap2}.  The difference with \autoref{fig:trap} is
that both the \textit{main}() and \textit{handler}() methods increment \textit{count}.
This is not unlike the example we gave in \autoref{fig:inc}, except that only a single
process is involved now.  And, indeed, it suffers from a similar race condition; run
it through Harmony to see for yourself.  If the interrupt occurs after \textit{main}()
read \textit{count} (and thus still has value 0) but before \textit{main()} writes the
updated value~1, then the interrupt handler will also read value~0 and write value~1.
We say that the code in \autoref{fig:trap2} is not \textit{interrupt-safe} (as opposed
to not being \textit{thread-safe}).
\index{interrupt-safety}

You would be excused if you wanted to solve the problem using locks, similar to
\autoref{fig:incfixed}.  \autoref{fig:trap3} shows how one might go about this.
But locks are intended to solve synchronization issues between multiple processes.
If you run the code through Harmony, you will find that the code may not terminate.
The issue is that a process can only acquire a lock once.  If the interrupt happens
after \textit{main}() acquires the lock but before \textit{main}() releases it, the
\textit{handler}() method will block trying to acquire the lock, even though
it is executed the same process that holds the lock.

@@

It will often be necessary to write code that is both interrupt-safe \emph{and}
thread-safe.  As you might expect, this involves both managing locks and
interrupt levels.
To increment \textit{count}, the interrupt level must be \textit{True} and
\textit{countlock} must be held.
\autoref{fig:trap6} gives an example of how this might be done.
The important rule to remember is for a process to disable interrupts \emph{before}
attempting to acquire a lock.

Try moving \texttt{lock}() to the beginning of \texttt{increment}()
and \texttt{unlock}() to the end of \texttt{increment}() and see what happens.
While Harmony will only report one faulty run, this incorrect code can lead to
the assertion failing as well as processes getting blocked indefinitely.

\section*{Exercises}
\begin{problems}
\item Make methods \texttt{produce}() and \texttt{consume}() in \autoref{fig:boundedbuffer}
both thread-safe and interrupt-safe.
\end{problems}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\textit{lv} = $e$ & \textit{lv} is an lvalue and $e$ is an expression\\
\hline
\textit{lv} \texttt{[op]=} $e$ & \texttt{op} is one of \texttt{+}, \texttt{-}, \texttt{*}, \texttt{/}, 
\texttt{and}, and \texttt{or}\\
\hline
\texttt{pass} & do nothing\\
\hline
\texttt{del} \textit{lv} & delete\\
\hline
\texttt{assert $b$ [, $x$]} & $b$ is a Boolean expression.  Optionally report value of expression $e$\\
\hline
\texttt{const v = $e$} & \texttt{v} is an identifier, $e$ is a constant expression\\
\hline
\texttt{def m($v$, ...): S} & \texttt{m} and \texttt{v} are identifiers, \texttt{S} a list of statements\\
\hline
\texttt{let $v$ = $e$, ...: S} & $v$ is an identifier, $e$ is an expression, \texttt{S} a list of statements\\
\hline
\texttt{if $b$: S} & $b$ is a Boolean expression, \texttt{S} a list of statements\\
\hline
\texttt{while $b$: S} & $b$ is a Boolean expression, \texttt{S} a list of statements\\
\hline
\texttt{for $v$ in $e$: S} & $v$ is an identifier, $e$ is a set evaluated in order,
                            \texttt{S} a list of statements\\
\hline
\texttt{atomic: S} & \texttt{S} a list of statements\\
\hline
\texttt{spawn m $e$ [, $t$]} & \texttt{m} is a method,
$e$ is an expression, $t$ is a tag (an expression) \\
\hline
\texttt{go $c$ $e$} & $c$ is a context, $e$ is an expression \\
\hline
\texttt{import m} & \texttt{m} identifies a module \\
\hline
\end{tabular}
\vspace{1em}

>>>>

Processes can be \emph{interrupted}.  An interrupt is a notification of some event
such as a keystroke, a timer expiring, the reception of a network packet, the completion
of a disk operation, and so on.  We distinguish \emph{interrupts} and \emph{exceptions}.
An exception is caused by the process executing an invalid machine instruction such as
divide-by-zero.  An interrupt is caused by some peripheral device and can be handled in
Harmony; an exception is an error.
\index{exception}

Harmony allows modeling interrupts using the \texttt{trap} statement:
\begin{code}
\textbf{trap} \textit{handler} \textit{argument}
\end{code}
invokes \textit{handler argument} at some later, unspecified time.
Only one of these asynchronous events can be outstanding at a time; a new call to
\textbf{trap} overwrites any outstanding one.
\autoref{fig:trap} gives an example of how \textbf{trap} might be used.
Here, the \textit{main}() process loops until the interrupt has occurred and
the \textit{done} flag has been set.

But now consider \autoref{fig:trap2}.  The difference with \autoref{fig:trap} is
that both the \textit{main}() and \textit{handler}() methods increment \textit{count}.
This is not unlike the example we gave in \autoref{fig:inc}, except that only a single
process is involved now.  And, indeed, it suffers from a similar race condition; run
it through Harmony to see for yourself.  If the interrupt occurs after \textit{main}()
reads \textit{count} (and thus still has value 0) but before \textit{main()} writes the
updated value~1, then the interrupt handler will also read value~0 and write value~1.
We say that the code in \autoref{fig:trap2} is not \textit{interrupt-safe} (as opposed
to not being \textit{thread-safe}).
\index{interrupt-safety}

You would be excused if you wanted to solve the problem using locks, similar to
\autoref{fig:incfixed}.  \autoref{fig:trap3} shows how one might go about this.
But locks are intended to solve synchronization issues between multiple processes.
If you run the code through Harmony, you will find that the code may not terminate.
The issue is that a process can only acquire a lock once.  If the interrupt happens
after \textit{main}() acquires the lock but before \textit{main}() releases it, the
\textit{handler}() method will block trying to acquire the lock, even though
it is being acquired by the same process that already holds the lock.

@@

It will often be necessary to write code that is both interrupt-safe \emph{and}
thread-safe.  As you might expect, this involves both managing locks and
interrupt levels.
To increment \textit{count}, the interrupt level must be \textit{True} and
\textit{countlock} must be held.
\autoref{fig:trap6} gives an example of how this might be done.
One important rule to remember is for a process to disable interrupts \emph{before}
attempting to acquire a lock.

Try moving \texttt{lock}() to the beginning of the \texttt{increment} method
and \texttt{unlock}() to the end of \texttt{increment} and see what happens.
While Harmony will only report one faulty run, this incorrect code can lead to
the assertion failing as well as processes getting blocked indefinitely.

There is another important rule to keep in mind.  Just locks should never be held
for long, interrupts should never be disabled for long.  With locks the issue is to
maximize concurrent performance.  The interrupts the issue is fast response to
asynchronous events.  Because interrupts may be disabled only briefly, interrupt
handlers must run quickly and cannot wait for other events.  So it is ok to invoke
synchronization calls such as \texttt{V} and \texttt{notify}, but calls such as
\texttt{P} and \texttt{wait} should only be used if it's known that they will not
block for long.  Informally, interrupt handlers can be
\emph{producers} but not \emph{consumers} of synchronization events.

\section*{Exercises}
\begin{problems}
\item The \texttt{produce} method in \autoref{fig:boundedbuffer} cannot be used
from interrupt handlers for two reasons: (1) it is not interrupt-safe, and (2)
it may block for a long time if the buffer is full.  Yet, it would be useful if,
say, a keyboard interrupt handler could place an event on a shared queue.  Implement
a new method \texttt{iproduce}(\textit{item}) that is interrupt-safe and does not
block.  Instead, it should return \texttt{False} if the buffer is full and \texttt{True}
if the item was successfully enqueued.
\end{problems}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\textit{lv} = $e$ & \textit{lv} is an lvalue and $e$ is an expression\\
\hline
\textit{lv} \texttt{[op]=} $e$ & \texttt{op} is one of \texttt{+}, \texttt{-}, \texttt{*}, \texttt{/}, 
\texttt{and}, and \texttt{or}\\
\hline
\texttt{pass} & do nothing\\
\hline
\texttt{del} \textit{lv} & delete\\
\hline
\texttt{assert $b$ [, $x$]} & $b$ is a Boolean expression.  Optionally report value of expression $e$\\
\hline
\texttt{const v = $e$} & \texttt{v} is an identifier, $e$ is a constant expression\\
\hline
\texttt{def m($v$, ...): S} & \texttt{m} and \texttt{v} are identifiers, \texttt{S} a list of statements\\
\hline
\texttt{let $v$ = $e$, ...: S} & $v$ is an identifier, $e$ is an expression, \texttt{S} a list of statements\\
\hline
\texttt{if $b$: S} & $b$ is a Boolean expression, \texttt{S} a list of statements\\
\hline
\texttt{while $b$: S} & $b$ is a Boolean expression, \texttt{S} a list of statements\\
\hline
\texttt{for $v$ in $e$: S} & $v$ is an identifier, $e$ is a set evaluated in order,
                            \texttt{S} a list of statements\\
\hline
\texttt{atomic: S} & \texttt{S} a list of statements\\
\hline
\texttt{spawn m $e$ [, $t$]} & \texttt{m} is a method,
$e$ is an expression, $t$ is a tag (an expression) \\
\hline
\texttt{trap m $e$} & \texttt{m} is a method and $e$ is an expression \\
\hline
\texttt{go $c$ $e$} & $c$ is a context, $e$ is an expression \\
\hline
\texttt{import m} & \texttt{m} identifies a module \\
\hline
\end{tabular}
\vspace{1em}
